Variance reduction for MCMC methods via martingale representations

### 2020-03-28 aap/Revision/Response_1.tex

\documentclass{article}%
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[authoryear]{natbib}
\usepackage{color}%
\setcounter{MaxMatrixCols}{30}
%TCIDATA{OutputFilter=latex2.dll}
%TCIDATA{Version=5.50.0.2890}
%TCIDATA{CSTFile=40 LaTeX article.cst}
%TCIDATA{Created=Thursday, September 14, 2017 13:22:20}
%TCIDATA{LastRevised=Saturday, December 07, 2019 12:32:10}
%TCIDATA{<META NAME="GraphicsSave" CONTENT="32">}
%TCIDATA{<META NAME="SaveForMode" CONTENT="1">}
%TCIDATA{BibliographyScheme=Manual}
%TCIDATA{<META NAME="DocumentShell" CONTENT="Standard LaTeX\Blank - Standard LaTeX Article">}
%BeginMSIPreambleData
\providecommand{\U}[1]{\protect\rule{.1in}{.1in}}
%EndMSIPreambleData
\newtheorem{theorem}{Theorem}
\newtheorem{acknowledgement}[theorem]{Acknowledgement}
\newtheorem{algorithm}[theorem]{Algorithm}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{case}[theorem]{Case}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{conclusion}[theorem]{Conclusion}
\newtheorem{condition}[theorem]{Condition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{criterion}[theorem]{Criterion}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{solution}[theorem]{Solution}
\newtheorem{summary}[theorem]{Summary}
\newenvironment{proof}[1][Proof]{\noindent\textbf{#1.} }{\ \rule{0.5em}{0.5em}}
\begin{document}

\title{\textbf{Response to the Report by Referee 1}}
\author{{\normalsize \vspace{-1cm} Denis Belomestny, Eric Moulines, and Sergei Samsonov}}
\date{~ }
\maketitle

\vspace{-0.5cm}

We thank the referee for carefully reading our manuscript and for pertinent
comments which helped us to improve the paper. In the new version of our
paper, we have taken all comments and suggestion into account.  Below are some additional responses to the
comments by Referee 1

\section*{Main remarks}

\begin{itemize}

\item \textit{... The authors make an attempt to quantify the variance reduction obtained by the proposed approach, as well as the computational complexity to achieve this variance reduction. On these topics several results are presented as Theorems and Propositions  ... I would recommend the authors to phrase this results with clear conditions in such a way that it becomes directly accessible to the reader ... where the bound is phrased in an easily readible way involving constants that are defined in the statement of the theorem and which should not depend on infinite series and conditional expectations.}
\par
Let us stress that  current literature on MCMC  mainly focuses on  convergence of MCMC algorithms in Total Variation and Wasserstein distances thus  ignoring a sample error of these algorithms. Only few references studied MSE error of the Langevin algorithm (see e.g.  Durmus and  Moulines (2017)). In this work we propose a novel variance reduction approach which is able to reduce the statistical error of MCMC algorithms. The main advantage of our method is provable increase of efficiency for Langevin-based MCMC algorithms as measured by  cost-times-variance criterion. Our approach is based on a new discrete time martingale representation for Markov chains which is of its own interest. Therefore we first present this basic result and then discuss in Section 4 variance reduction method based on the derived martingale representation. In  Section~4 we  explain in accessible way why our approach leads to reduced cost-times-variance criterion. These results are based on deep analysis of the variance reduced estimate postponed to Section 5. In the revised version we significantly weaken  assumptions and prove our main result (Theorem~3 and Corollary~2) for ULA under smoothness  and convexity of the underlying potential outside a ball (assumption  (H2)).   Note that all assumptions needed are now collected in Theorem~3. For convenience of the reader we also sketch the main steps of the proof right after 
Theorem~3. We decided to present our main theoretical results as bounds for the functions $A_{q,k}$ since these functions play a crucial role in the variance bound (13). Given these estimates it is easy to see how to get bounds for the variance. 
\item \textit{... It is good to see that the technique has beneficial effects on the variance, but it does not become clear what was the computational effort required to achieve this benefit. The reader is left wondering whether or not the same variance reduction could have been achieved simply by drawing more MCMC samples ...}
\item \textit{... here the setting \(\gamma_k = c k^{-\alpha}, \alpha\in  (0, 1)\) is considered, and presented as  ``the most interesting case''. For many practitioners, $\alpha=0$ would be rather natural. What can be said in this important case? This is also applies to Section 6 ...}
\par
In the revised version we decided to switch to a constant time step and present all our bounds for this case. On the one hand, this allows us to formulate our main results in a simpler form depending only on a single parameter $\gamma.$ On the other hand, this is indeed the most relevant choice for practitioners.  
\item \textit{ ... here the functions g are assumed uniformly bounded, whereas earlier (end of Section 5) elements in $\Psi$ were proposed to be polynomials. In particular the bound B would be equal to infinity. How can this be reconciled? ...}
\par
In the revised version we skip this result because of size restrictions. 
\end{itemize}

\end{document}

### 2020-03-28 aap/Revision/Response_2.tex

\documentclass{article}%
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[authoryear]{natbib}
\usepackage{color}%
\setcounter{MaxMatrixCols}{30}
%TCIDATA{OutputFilter=latex2.dll}
%TCIDATA{Version=5.50.0.2890}
%TCIDATA{CSTFile=40 LaTeX article.cst}
%TCIDATA{Created=Thursday, September 14, 2017 13:22:20}
%TCIDATA{LastRevised=Saturday, December 07, 2019 12:32:10}
%TCIDATA{<META NAME="GraphicsSave" CONTENT="32">}
%TCIDATA{<META NAME="SaveForMode" CONTENT="1">}
%TCIDATA{BibliographyScheme=Manual}
%TCIDATA{<META NAME="DocumentShell" CONTENT="Standard LaTeX\Blank - Standard LaTeX Article">}
%BeginMSIPreambleData
\providecommand{\U}[1]{\protect\rule{.1in}{.1in}}
%EndMSIPreambleData
\newtheorem{theorem}{Theorem}
\newtheorem{acknowledgement}[theorem]{Acknowledgement}
\newtheorem{algorithm}[theorem]{Algorithm}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{case}[theorem]{Case}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{conclusion}[theorem]{Conclusion}
\newtheorem{condition}[theorem]{Condition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{criterion}[theorem]{Criterion}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{solution}[theorem]{Solution}
\newtheorem{summary}[theorem]{Summary}
\newenvironment{proof}[1][Proof]{\noindent\textbf{#1.} }{\ \rule{0.5em}{0.5em}}
\begin{document}

\title{\textbf{Response to the Report by Referee 2}}
\author{{\normalsize \vspace{-1cm} Denis Belomestny, Eric Moulines, and Sergei Samsonov}}
\date{~ }
\maketitle

\vspace{-0.5cm}

We thank the referee for carefully reading our manuscript and for pertinent
comments which helped us to improve the paper. In the new version of our
paper, we have taken all comments and suggestion into account.  Below are some additional responses to the
comments by Referee 2

@@

\section*{Main remarks}

\begin{itemize}

\item \textit{... Restrictive assumptions for the theoretical analysis. The main theorems of the paper, which seem to be Theorem 4 and the results of section 6.1, have the restrictive assumption that the drift vector field  needs to satisfy the strong convexity assumption ... In typically interesting cases, such as Bayesian inverse problems and machine learning, this assumption is not satisfied. Unfortunately, even in the numerical examples provided in Section 7, this assumption is not fulfilled ? it already doesn't hold for a two-component Gaussian mixture in one dimension. This is a serious restriction ... }
\par
In the revised version we significantly weaken  the assumptions and prove our main result (Theorem~3 and Corollary~2) for ULA under smoothness  and convexity of the underlying potential outside a ball (assumption  (H2)).  Thus we avoid the strong convexity assumption.  Note that all assumptions needed are now collected in Theorem~3. These  assumptions can now be validated for a large class of examples including example of Gaussian mixtures considered in Section~6.1.
\item \textit{... in complicated models such as PDE-related Bayesian inverse problems or machine learning, the computation of \(\nabla \log(\pi)\) is expensive ? but these examples are typically ones where \(\log(\pi)\) does not satisfy the needed concavity assumptions ...}
\par
Now we replaced the strong convexity assumption by a much weaker one (H2) (convexity outside a ball). Hence our method may gain additional efficiency as compared to other variance reduction approaches due to the fact that \(\nabla \log(\pi)\) doesn't need to be computed to construct control variates.
\end{itemize}

### 2020-04-09 aap/Revision/vr-mcmc-revision-new-new.tex
update

= \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{k_i,s_i}(x_i),

@@

\widehat M_{K,n,n_{\mathrm{trunc}}}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{0 < \|\mathbf{k}\| < K}\sum_{l=N+1}^{p} \widehat a_{p-l,\mathbf{k}}(X_{l-1})\phi_k(\xi_{l})\mathbbm{1}\{ | p-l | < n_{\mathrm{trunc}}\right].

@@

\subsection{Gaussian mixtures}
 We consider a sample generated by ULA with $\pi$  given by the mixture of two equally-weighted $d$-dimensional Gaussian distributions of the following form
 \[
 \pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}|\Sigma|}} \left( \rme^{-\frac{(x-\mu)^T\Sigma^{-1}(x-\mu)}{2}} + \rme^{-\frac{(x+\mu)^T\Sigma^{-1}(x+\mu)}{2}}\right)
 \]
where $\rho \in (0,1), \mu \in \rset^d, \Sigma \in \rset^{d \times d}$ and $|\Sigma|$ is its determinant. The function $U(x)$ and its gradient are given by
\[
U(x) = \frac{1}{2}(x-\mu)^T\Sigma^{-1}(x-\mu) - \ln{\left(1 + \rme^{-2\mu^\top \Sigma^{-1}x}\right)}, \quad
\]
\[
\nabla U(x) = \Sigma^{-1}(x-\mu) +2\mu^\top\Sigma^{-1}\left(1 + \rme^{2 \mu^\top\Sigma^{-1} x}\right)^{-1},
\]
respectively.
In our experiments we considered $d = 2$, $\mu = \left(0.5,0.5\right)$ and randomly generated positive-definite matrix $\Sigma$ (heterogeneous structure). In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i\) or \(f(x) = \sum_{i=1}^d x^2_i\) we used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 10$ independent training trajectories, each one of size $n = 10^4$.  Then we solve the least squares  problems \eqref{eq:Qregr} using the first order polynomial approximations for the coefficients \(\hat{a}_{p,\mathbf{k}}\) as described in the previous section, hence we set $K = 1$. We set the truncation level \(n_{\mathrm{trunc}} = 50\). To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 10^3$ and plot boxplots of the ergodic averages in Figure ~\ref{fig:GMM}. Our approach (MDCV, Martingale decomposition control variates) is compared to other variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}. In the baselines we use first order polynomials, since they are used as regressors during computations of $\hat{a}_{p-l,\mathbf{k}}$.

 \begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/GMM_2d_lin_sum.png}
\label{fig:subfig1}}
\qquad
\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/GMM_2d_lin_sum_squares.png}
\label{fig:subfig2}}
\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Gaussian mixture model. The compared estimators are the ordinary empirical average  (Vanilla), our estimator of control variates (MDCV), zero variance estimator (ZV) and control variates obtained with empirical spectral variance minimisation (ESVM) 
\label{fig:GMM}}
\end{figure}

@@

\section{Covariance estimation for $V-$geometrically ergodic Markov chains}\label{sec:appendix_moments}
In this section we assume that $(X_p)_{p \in \nset_0}$ is a $V-$geometrically ergodic Markov chain and prove bounds on the variance of ergodic average $\pi_N^n(f)$ of the form \eqref{eq:29032018a2}. We use the same technique as in \cite{belomestny2019esvm} to control autocovariances for a given Markov chain. We start from auxiliary lemma:
\begin{lem} 
\label{lem:fp_covariance}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Then it holds for some constant $C > 0$
\begin{equation}
|\mathsf{E}_x[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq C\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x)
\end{equation}
For the stationary distribution it holds
\begin{equation}
|\mathsf{E}_{\pi}[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq C\pi(V)\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}
\end{equation}
\end{lem}
\begin{proof}
Note that
\[
|\mathsf{E}_x[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq |\tilde{f}(x)| \mathsf{E}_x|\tilde{f}(X_s)| \leq \|\tilde{f}\|_{V^{\frac{1}{2}}}V^{\frac{1}{2}}(x)\int\limits_{\rset^d} V^{\frac{1}{2}}(y)|\mathsf{P}^{s}(x,\,dy)-\pi(dy)|
\]
By Hoelder inequality, 
\begin{eqnarray*}
\int\limits_{\rset^d} V^{\frac{1}{2}}(y)|\mathsf{P}^{s}(x,\,dy)-\pi(dy)| & \leq & \left(\int\limits_{\rset^d} V(y)|\mathsf{P}^{s}(x,\,dy)-\pi(dy)|\right)^{1/2}\left(\int\limits_{\rset^d}|\mathsf{P}^{s}(x,\,dy)-\pi(dy)|\right)^{1/2} \leq \\
& \leq & 2\rho^{s/2}V^{1/2}(x)
\end{eqnarray*}
yielding the first statement of lemma. The second statement can be obtained from the first one by integration with respect to $\pi$. 
\end{proof}

\begin{lem}
\label{lem:covariance}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Then
\begin{equation}
\label{eq:cov_v_ergodic}
\PCov_x \left[f(X_k),f(X_{k+s})\right] \leq C^2V^2(x)\rho^{k+s} + C^2\rho^{k+s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x) + C\pi(V)\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}
\end{equation}
\end{lem}
\begin{proof} 
Note that
\[
\PCov_x \left[f(X_k),f(X_{k+s})\right] = \PE_x\left[f(X_k) - \pi(f)\right]\left[f(X_{k+s})-\pi(f)\right] + \left[\pi(f) - \PE_xf(X_s)\right]\left[\PE_xf(X_k) - \pi(f)\right]
\]
Due to $V-$ergodicity, 
\[
\left|\left[\pi(f) - \PE_xf(X_s)\right]\left[\PE_xf(X_k) - \pi(f)\right]\right| \leq C^2V^2(x)\rho^{k+s}
\]
To bound the first term note that
\begin{multline*}
\left|\PE_x\left[f(X_k) - \pi(f)\right]\left[f(X_{k+s})-\pi(f)\right] - \PCov_{\pi}\left[f(X_k),f(X_{k+s})\right]\right| 
\\
\leq \int\limits_{\rset^d}\left|\PE_y\left[\tilde{f}(X_0)\tilde{f}(X_{s})\right]\right| |\P^k(x,dy)-\pi(dy)|  
\leq C^2\rho^{k+s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x),
\end{multline*}
where the last inequality is due to lemma~\ref{lem:fp_covariance}, which implies \eqref{eq:cov_v_ergodic}.
\end{proof}
Now we state and prove the main result of this section on the variance bound for the estimator $\pi_N^n(f)$ for $V-$geometrically ergodic Markov chain.
\begin{lem}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Assume also that $n(1-\rho^{1/2}) > 1$. Then 
\begin{equation}
\label{eq:vanilla_var}
\PVar_x \left[ \pi_{N}^n(f) \right] \leq \frac{C\pi(V)\|\tilde{f}\|_{V^{1/2}}}{n(1-\rho^{1/2})} + \mathcal{O}(\frac{1}{n^2(1-\rho^{1/2})}) 
\end{equation}
\end{lem}
\begin{proof}
Note that 
\begin{eqnarray*}
\PVar_x \left[ \frac{1}{n}\sum\limits_{k=N+1}^{N+n}f(X_k) \right] = \underbrace{\frac{1}{n^2}\sum\limits_{k=N+1}^{N+n}\PVar_x \left[ f(X_k) \right]}_{S_1} + \underbrace{\frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\sum\limits_{s=1}^{n-k-1}\PCov_x\left[f(X_k),f(X_{k+s})\right]}_{S_2}
\end{eqnarray*}
Now we bound first and second sum using lemma~\ref{lem:covariance}:
\[
S_1 \leq \frac{1}{n}C\pi(V)\|\tilde{f}\|_{V^{1/2}} + \frac{\rho^N C^2\left(V^2(x) + V(x)\|\tilde{f}\|_{V^{1/2}}\right)}{n^2(1-\rho)}
\]
\begin{eqnarray*}
S_2 & \leq & \frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\left[C^2V^2(x)\rho^{k+1}\frac{1}{1-\rho} + C^2V(x)\rho^{k+1/2}\frac{1}{1-\rho^{1/2}} + C\pi(V)\frac{1}{1-\rho^{1/2}}\|\tilde{f}\|_{V^{1/2}}\right] \leq \\
& \leq & \frac{2\rho^NC^2V^2(x)}{n^2(1-\rho)^2} + \frac{2\rho^NC^2V(x)}{n^2(1-\rho)(1-\rho^{1/2})} + \frac{C\pi(V)\|\tilde{f}\|_{V^{1/2}}}{n(1-\rho^{1/2})}
\end{eqnarray*}
Hence \eqref{eq:vanilla_var} follows.
\end{proof}

>>>>

&=& 
\label{eq:a-pol}
\sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{k_i,s_i}(x_i - \gamma \mu_i(x)),

@@

\widehat M_{K,n,n_{\mathrm{trunc}}}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{0 < \|\mathbf{k}\|\leq K}\sum_{l=N+1}^{p} \widehat a_{p-l,\mathbf{k}}(X_{l-1})\mathbf{H}_\mathbf{k}(\xi_{l})\mathbbm{1}\{ | p-l | < n_{\mathrm{trunc}}\}\right].

@@

\subsection{Gaussian mixtures}
 We consider  ULA with $\pi$  given by the mixture of two equally-weighted $2$-dimensional Gaussian distributions of the following form
\begin{eqnarray}
\label{eq:gmm}
\pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}|\Sigma|}} \left( \rme^{-\frac{(x-\mu)^T\Sigma^{-1}(x-\mu)}{2}} + \rme^{-\frac{(x+\mu)^T\Sigma^{-1}(x+\mu)}{2}}\right)
\end{eqnarray}
where $\rho \in (0,1), \mu \in \rset^d, \Sigma \in \rset^{2 \times 2}$ and $|\Sigma|$ is its determinant. The function $U(x)$ and its gradient are given by
\[
U(x) = \frac{1}{2}(x-\mu)^T\Sigma^{-1}(x-\mu) - \ln{\left(1 + \rme^{-2\mu^\top \Sigma^{-1}x}\right)}
\]
and
\[
\nabla U(x) = \Sigma^{-1}(x-\mu) +2\mu^\top\Sigma^{-1}\left(1 + \rme^{2 \mu^\top\Sigma^{-1} x}\right)^{-1},
\]
respectively.
In our experiments we take $\mu = \left(0.5,0.5\right)$ and randomly generated positive-definite matrix $\Sigma$ (heterogeneous structure). In order to approximate the expectation \(\pi(f)\) for \(f(x)=\sum_{i=1}^d x_i\) or \(f(x) = \sum_{i=1}^d x^2_i,\) we use ULA with constant step size $\gamma=0.1$ and sampled $\NtrainPath = 10$ independent training trajectories, each one of size $n = 10^4$ to estimate coefficients $a_{p-l,\mathbf{k}}.$ Then we solve the least squares  problems \eqref{eq:Qregr} using the first or second order polynomial basis  depending on $f$. Respectively, we set $K=1$ for linear functionals $f$ and $K=2$ for quadratic functionals. We set the truncation level \(n_{\mathrm{trunc}} = 50\). To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 10^3$ and depict boxplots of the ergodic averages in Figure ~\ref{fig:GMM}. Our approach (MDCV, Martingale decomposition control variates) is compared to other variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}. In the baselines we use first order polynomials, as they were used as regressors during computations of $\hat{a}_{p-l,\mathbf{k}}$. Next we compute the cost-variance ratios
\[
\mathcal{R}(K,N,n,n_{\mathrm{trunc}})=\frac{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n,n_{\mathrm{trunc}} }^N(f)|\TrainSet]} 
\]
where variances in the above definition are computed using $24$ independent trajectories of the length $n=2\cdot10^3.$ Note that the cost of $\widehat \pi_{K,n,n_{\mathrm{trunc}}} $ is proportional to the number of nonzero coefficients in \eqref{eq:a-pol} times $n_{\mathrm{trunc}}$. In Figure~\ref{fig:G-costvar} cost-variances ratios are shown for for the case of  Gaussian distribution with parameters. In particular, in Figure~\ref{fig:subfig3}  a function $\mathcal{R}(1,2\cdot 10^3,2\cdot 10^3,n_{\mathrm{trunc}})$ is shown in dependence on $n_{\mathrm{trunc}}$ for the case of a linear functional $f$. In Figure~\ref{fig:subfig4} a similar plot of $\mathcal{R}(2,2\cdot 10^3,2\cdot 10^3,n_{\mathrm{trunc}})$ is shown for the case of a quadratic functional $f.$
Figure~\ref{fig:GMM-costvar} presents the corresponding coast-variance ratios   for a mixture of two two-dimensional normal distributions as in \eqref{eq:gmm} with $\Sigma=I$ and $\mu=(0.5,0.5).$ 
 \begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/GMM_2d_lin_sum.png}
\label{fig:subfig1}}
\qquad
\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/GMM_2d_lin_sum_squares.png}
\label{fig:subfig2}}
\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Gaussian mixture model. The compared estimators are the ordinary empirical average  (Vanilla), our estimator of control variates (MDCV), zero variance estimator (ZV) and control variates obtained with empirical spectral variance minimisation (ESVM) 
\label{fig:GMM}}
\end{figure}

\begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_linear_gaussian.png}
\label{fig:subfig3}}
\qquad
\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_quadratic_gaussian.png}
\label{fig:subfig4}}
\caption{ Cost-variance ratios as a function of the truncation level for  two-dimensional standard Gaussian distribution and different test functions. \label{fig:G-costvar}}
\end{figure}

\begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_linear_gmm.png}
\label{fig:subfig5}}
\qquad
\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_quadratic_gmm.png}
\label{fig:subfig6}}
\caption{ Cost-variance ratios as a function of the truncation level for  a mixture of two-dimensional  Gaussian distribution and different test functions.
\label{fig:GMM-costvar}}
\end{figure}

@@

\section{Covariance estimation for $V-$geometrically ergodic Markov chains}\label{sec:appendix_moments}
In this section we assume that $(X_p)_{p \in \nset_0}$ is a $V-$geometrically ergodic Markov chain and prove bounds on the variance of ergodic average $\pi_N^n(f)$ of the form \eqref{eq:29032018a2}. We use the same technique as in \cite{belomestny2019esvm} to control autocovariances for a given Markov chain. We start from auxiliary lemma:
\begin{lem} 
\label{lem:fp_covariance}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Let $\tilde{f}(x) = f(x) - \pi(f)$. Then it holds for some constant $C > 0$
\begin{equation}
|\mathsf{E}_x[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq C\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x)
\end{equation}
For the stationary distribution it holds
\begin{equation}
|\mathsf{E}_{\pi}[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq C\pi(V)\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}
\end{equation}
\end{lem}
\begin{proof}
Note that
\[
|\mathsf{E}_x[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq |\tilde{f}(x)| \mathsf{E}_x|\tilde{f}(X_s)| \leq \|\tilde{f}\|_{V^{\frac{1}{2}}}V^{\frac{1}{2}}(x)\int\limits_{\rset^d} V^{\frac{1}{2}}(y)|\mathsf{P}^{s}(x,\,dy)-\pi(dy)|
\]
By Hoelder inequality, 
\begin{eqnarray*}
\int\limits_{\rset^d} V^{\frac{1}{2}}(y)|\mathsf{P}^{s}(x,\,dy)-\pi(dy)| & \leq & \left(\int\limits_{\rset^d} V(y)|\mathsf{P}^{s}(x,\,dy)-\pi(dy)|\right)^{1/2}\left(\int\limits_{\rset^d}|\mathsf{P}^{s}(x,\,dy)-\pi(dy)|\right)^{1/2} \leq \\
& \leq & 2\rho^{s/2}V^{1/2}(x)
\end{eqnarray*}
yielding the first statement of lemma. The second statement can be obtained from the first one by integration with respect to $\pi$. 
\end{proof}

\begin{lem}
\label{lem:covariance}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Let $\tilde{f}(x) = f(x) - \pi(f)$. Then
\begin{equation}
\label{eq:cov_v_ergodic}
\PCov_x \left[f(X_k),f(X_{k+s})\right] \leq C^2V^2(x)\rho^{k+s} + C^2\rho^{k+s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x) + C\pi(V)\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}
\end{equation}
\end{lem}
\begin{proof} 
Note that
\[
\PCov_x \left[f(X_k),f(X_{k+s})\right] = \PE_x\left[f(X_k) - \pi(f)\right]\left[f(X_{k+s})-\pi(f)\right] + \left[\pi(f) - \PE_xf(X_s)\right]\left[\PE_xf(X_k) - \pi(f)\right]
\]
Due to $V-$ergodicity, 
\[
\left|\left[\pi(f) - \PE_xf(X_s)\right]\left[\PE_xf(X_k) - \pi(f)\right]\right| \leq C^2V^2(x)\rho^{k+s}
\]
To bound the first term note that
\begin{multline*}
\left|\PE_x\left[f(X_k) - \pi(f)\right]\left[f(X_{k+s})-\pi(f)\right] - \PCov_{\pi}\left[f(X_k),f(X_{k+s})\right]\right| 
\\
\leq \int\limits_{\rset^d}\left|\PE_y\left[\tilde{f}(X_0)\tilde{f}(X_{s})\right]\right| |\P^k(x,dy)-\pi(dy)|  
\leq C^2\rho^{k+s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x),
\end{multline*}
where the last inequality is due to lemma~\ref{lem:fp_covariance}, which implies \eqref{eq:cov_v_ergodic}.
\end{proof}
Now we state and prove the main result of this section on the variance bound for the estimator $\pi_N^n(f)$ for $V-$geometrically ergodic Markov chain.
\begin{lem}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Assume also that $n(1-\rho^{1/2}) > 1$. Then 
\begin{equation}
\label{eq:vanilla_var}
\PVar_x \left[ \pi_{N}^n(f) \right] \leq \frac{C\pi(V)\|\tilde{f}\|_{V^{1/2}}}{n(1-\rho^{1/2})} + \mathcal{O}(\frac{1}{n^2(1-\rho^{1/2})}) 
\end{equation}
\end{lem}
\begin{proof}
Note that 
\begin{eqnarray*}
\PVar_x \left[ \frac{1}{n}\sum\limits_{k=N+1}^{N+n}f(X_k) \right] = \underbrace{\frac{1}{n^2}\sum\limits_{k=N+1}^{N+n}\PVar_x \left[ f(X_k) \right]}_{S_1} + \underbrace{\frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\sum\limits_{s=1}^{n-k-1}\PCov_x\left[f(X_k),f(X_{k+s})\right]}_{S_2}
\end{eqnarray*}
Now we bound first and second sum using lemma~\ref{lem:covariance}:
\[
S_1 \leq \frac{1}{n}C\pi(V)\|\tilde{f}\|_{V^{1/2}} + \frac{\rho^N C^2\left(V^2(x) + V(x)\|\tilde{f}\|_{V^{1/2}}\right)}{n^2(1-\rho)}
\]
\begin{eqnarray*}
S_2 & \leq & \frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\left[C^2V^2(x)\rho^{k+1}\frac{1}{1-\rho} + C^2V(x)\rho^{k+1/2}\frac{1}{1-\rho^{1/2}} + C\pi(V)\frac{1}{1-\rho^{1/2}}\|\tilde{f}\|_{V^{1/2}}\right] \leq \\
& \leq & \frac{2\rho^NC^2V^2(x)}{n^2(1-\rho)^2} + \frac{2\rho^NC^2V(x)}{n^2(1-\rho)(1-\rho^{1/2})} + \frac{C\pi(V)\|\tilde{f}\|_{V^{1/2}}}{n(1-\rho^{1/2})}
\end{eqnarray*}
Hence \eqref{eq:vanilla_var} follows.
\end{proof}
